scRNAseq

### 2018-02-27 datasets/scrnaseq/scRNAseq-keegan.Rmd
Organize sc analyses into one dataset/method per Rmd

Here we are interested in performing differential expression analyses to determine
which

@@

cores <- 8
BiocParallel::register(MulticoreParam(workers = cores))

@@

Per Soneson et al., we run TMM normalization to the TPM units before running Wilcoxon.

@@

We apply scDD to normalized log-transformed counts.

@@

## Covariate Diagnostics
** Still need to clean up from here on; nothing below is eval'd**

In scrnaseq data, strength of the signal can be a function of level of gene expression. We quantify it as median or mean gene expression,and detection rate, defined as the percentage of cells expressing the gene at a level of TPM>0.

```{r,cov, eval=FALSE}
h_cov=compute_covariates_scrnaseq(paste0(datdir,"h_g_nofilter_subdata_GSE63818-GPL16791.rds"))
m_cov=compute_covariates_scrnaseq(paste0(datdir,"m_g_nofilter_subdata_GSE45719.rds"))
h2_cov=compute_covariates_scrnaseq(paste0(datdir,"h_g_nofilter_subdata_GSE84465.rds"))
m2_cov=compute_covariates_scrnaseq(paste0(datdir,"m_g_nofilter_subdata_GSE94383.rds"))
```

First we will look at the covariates for the Wilcoxon text

```{r, wilcoxcov, fig.width=10, fig.height=7, eval=FALSE}
print("human_plate_1")
hwilcox1 <- plot_covariate_diagnostics(paste0(h_wilcox_dat1),"wilcox",h_cov)
print(hwilcox1$plot)

print("human_plate_2")
hwilcox2 <- plot_covariate_diagnostics(paste0(h_wilcox_dat2),"wilcox",h2_cov)
print(hwilcox2$plot)

print("mouse_plate_1")
mwilcox1 <- plot_covariate_diagnostics(paste0(m_wilcox_dat1),"wilcox",m_cov)
print(mwilcox1$plot)

print("mouse_plate_2")
mwilcox2 <- plot_covariate_diagnostics(paste0(m_wilcox_dat2),"wilcox",m2_cov)
print(mwilcox2$plot)
```

Next, we look at the covariates for MAST

```{r, mastcov, fig.width=10, fig.height=7, eval=FALSE}
print("human_plate_1")
hmast1 <- plot_covariate_diagnostics(paste0(h_mast_dat1),"mast",h_cov)
print(hmast1$plot)

print("human_plate_2")
hmast2 <- plot_covariate_diagnostics(paste0(h_mast_dat2),"mast",h2_cov)
print(hmast2$plot)

print("mouse_plate_1")
mmast1 <- plot_covariate_diagnostics(paste0(m_mast_dat1),"mast",m_cov)
print(mmast1$plot)

print("mouse_plate_2")
mmast2 <- plot_covariate_diagnostics(paste0(m_mast_dat2),"mast",m2_cov)
print(mmast2$plot)
```

Finally, we look at how informative the covariates are with results from ScDD.

```{r,scddcov, fig.width=10, fig.height=7, eval=FALSE}
print("human_plate_1")
hscdd1 <- plot_covariate_diagnostics(paste0(h_scdd_dat1),"scdd",h_cov)
print(hscdd1$plot)

print("human_plate_2")
hscdd2 <- plot_covariate_diagnostics(paste0(h_scdd_dat2),"scdd",h2_cov)
print(hscdd2$plot)

print("mouse_plate_1")
mscdd1 <- plot_covariate_diagnostics(paste0(m_scdd_dat1),"scdd",m_cov)
print(mscdd1$plot)

print("mouse_plate_2")
mscdd2 <- plot_covariate_diagnostics(paste0(m_scdd_dat2),"scdd",m2_cov)
print(mscdd2$plot)
```

@@

First we will create the benchdesign object for wilcox -- IN PROGRESS

```{r, benchmarking, eval=FALSE, eval=FALSE}

>>>>

Here we are interested in performing differential expression analyses to determine
which genes have different expression among the two biological groups in each comparison

@@

cores <- 6
BiocParallel::register(MulticoreParam(workers = cores))

@@

Per Soneson et al., we run TMM normalization to the TPM units before running Wilcoxon.
The p-value results are added to the `rowData` slot of the `SummarizedExperiment` 
data objects to ensure that gene metadata stays intact.

@@

We apply scDD to normalized log-transformed counts. 
The p-value results are added to the `rowData` slot of the `SummarizedExperiment` 
data objects to ensure that gene metadata stays intact.

@@

## Covariate Diagnostics

In scrnaseq data, strength of the signal can be a affected by the of level of gene expression. 
We will explore two potential covariates related to expression level: mean expression 
and detection rate (defined as the proportion of cells expressing the gene at a nonzero level). 
These will also be added to the `rowData` slot of the `SummarizedExperiments`.

```{r,cov}
add_covariates_scrnaseq <- function(dat){
  rowData(dat)$meanExp <- rowMeans(assays(dat)[["count"]])
  rowData(dat)$detection <- rowMeans(assays(dat)[["count"]]==0)
  return(dat)
}

humandata <- add_covariates_scrnaseq(humandata)
humandata2 <- add_covariates_scrnaseq(humandata2)
mousedata <- add_covariates_scrnaseq(mousedata)
mousedata2 <- add_covariates_scrnaseq(mousedata2)

### Covariate one: Mean Expression

For each covariate, we'll examine the plots for each of the three analysis methods.

#### Wilcox

```{r, fig.width=4.5, fig.height=3.5, message=FALSE}
rank_scatter(data.frame(rowData(humandata)), 
             pvalue="wilcox_p", covariate="meanExp") + 
  ggtitle("Wilcox, Covariate 1: Mean Expression")
rank_scatter(data.frame(rowData(humandata2)), 
             pvalue="wilcox_p", covariate="meanExp") + 
  ggtitle("Wilcox, Covariate 1: Mean Expression")
rank_scatter(data.frame(rowData(mousedata)), 
             pvalue="wilcox_p", covariate="meanExp") + 
  ggtitle("Wilcox, Covariate 1: Mean Expression")
rank_scatter(data.frame(rowData(mousedata2)), 
             pvalue="wilcox_p", covariate="meanExp") + 
  ggtitle("Wilcox, Covariate 1: Mean Expression")
```

```{r, fig.width=10, fig.height=3.2, message=FALSE}
strat_hist(data.frame(rowData(humandata)), 
           pvalue="wilcox_p", covariate="meanExp", maxy=25) 
strat_hist(data.frame(rowData(humandata2)), 
           pvalue="wilcox_p", covariate="meanExp", maxy=25) 
strat_hist(data.frame(rowData(mousedata)), 
           pvalue="wilcox_p", covariate="meanExp", maxy=25) 
strat_hist(data.frame(rowData(mousedata2)), 
           pvalue="wilcox_p", covariate="meanExp", maxy=25) 
```

#### scDD

```{r, fig.width=4.5, fig.height=3.5, message=FALSE}
rank_scatter(data.frame(rowData(humandata)), 
             pvalue="scdd_p", covariate="meanExp") + 
  ggtitle("scDD, Covariate 1: Mean Expression")
rank_scatter(data.frame(rowData(humandata2)), 
             pvalue="scdd_p", covariate="meanExp") + 
  ggtitle("scDD, Covariate 1: Mean Expression")
rank_scatter(data.frame(rowData(mousedata)), 
             pvalue="scdd_p", covariate="meanExp") + 
  ggtitle("scDD, Covariate 1: Mean Expression")
rank_scatter(data.frame(rowData(mousedata2)), 
             pvalue="scdd_p", covariate="meanExp") + 
  ggtitle("scDD, Covariate 1: Mean Expression")
```

```{r, fig.width=10, fig.height=3.2, message=FALSE}
strat_hist(data.frame(rowData(humandata)), 
           pvalue="scdd_p", covariate="meanExp", maxy=25) 
strat_hist(data.frame(rowData(humandata2)), 
           pvalue="scdd_p", covariate="meanExp", maxy=25) 
strat_hist(data.frame(rowData(mousedata)), 
           pvalue="scdd_p", covariate="meanExp", maxy=25) 
strat_hist(data.frame(rowData(mousedata2)), 
           pvalue="scdd_p", covariate="meanExp", maxy=25) 
```

#### MAST

```{r, fig.width=4.5, fig.height=3.5, message=FALSE}
rank_scatter(data.frame(rowData(humandata)), 
             pvalue="mast_p", covariate="meanExp") + 
  ggtitle("MAST, Covariate 1: Mean Expression")
rank_scatter(data.frame(rowData(humandata2)), 
             pvalue="mast_p", covariate="meanExp") + 
  ggtitle("MAST, Covariate 1: Mean Expression")
rank_scatter(data.frame(rowData(mousedata)), 
             pvalue="mast_p", covariate="meanExp") + 
  ggtitle("MAST, Covariate 1: Mean Expression")
rank_scatter(data.frame(rowData(mousedata2)), 
             pvalue="mast_p", covariate="meanExp") + 
  ggtitle("MAST, Covariate 1: Mean Expression")
```

```{r, fig.width=10, fig.height=3.2, message=FALSE}
strat_hist(data.frame(rowData(humandata)), 
           pvalue="mast_p", covariate="meanExp", maxy=25) 
strat_hist(data.frame(rowData(humandata2)), 
           pvalue="mast_p", covariate="meanExp", maxy=25) 
strat_hist(data.frame(rowData(mousedata)), 
           pvalue="mast_p", covariate="meanExp", maxy=25) 
strat_hist(data.frame(rowData(mousedata2)), 
           pvalue="mast_p", covariate="meanExp", maxy=25) 
```

The mean expression appears to be informative and approximately satisfies the 
assumptions

### Covariate two: Detection Rate

Next we look at the detection rate covariate.

#### Wilcox

```{r, fig.width=4.5, fig.height=3.5, message=FALSE}
rank_scatter(data.frame(rowData(humandata)), 
             pvalue="wilcox_p", covariate="detection") + 
  ggtitle("Wilcox, Covariate 2: Detection Rate")
rank_scatter(data.frame(rowData(humandata2)), 
             pvalue="wilcox_p", covariate="detection") + 
  ggtitle("Wilcox, Covariate 2: Detection Rate")
rank_scatter(data.frame(rowData(mousedata)), 
             pvalue="wilcox_p", covariate="detection") + 
  ggtitle("Wilcox, Covariate 2: Detection Rate")
rank_scatter(data.frame(rowData(mousedata2)), 
             pvalue="wilcox_p", covariate="detection") + 
  ggtitle("Wilcox, Covariate 2: Detection Rate")
```

```{r, fig.width=10, fig.height=3.2, message=FALSE}
strat_hist(data.frame(rowData(humandata)), 
           pvalue="wilcox_p", covariate="detection", maxy=25) 
strat_hist(data.frame(rowData(humandata2)), 
           pvalue="wilcox_p", covariate="detection", maxy=25) 
strat_hist(data.frame(rowData(mousedata)), 
           pvalue="wilcox_p", covariate="detection", maxy=25) 
strat_hist(data.frame(rowData(mousedata2)), 
           pvalue="wilcox_p", covariate="detection", maxy=25) 
```

#### scDD

```{r, fig.width=4.5, fig.height=3.5, message=FALSE}
rank_scatter(data.frame(rowData(humandata)), 
             pvalue="scdd_p", covariate="detection") + 
  ggtitle("scDD, Covariate 2: Detection Rate")
rank_scatter(data.frame(rowData(humandata2)), 
             pvalue="scdd_p", covariate="detection") + 
  ggtitle("scDD, Covariate 2: Detection Rate")
rank_scatter(data.frame(rowData(mousedata)), 
             pvalue="scdd_p", covariate="detection") + 
  ggtitle("scDD, Covariate 2: Detection Rate")
rank_scatter(data.frame(rowData(mousedata2)), 
             pvalue="scdd_p", covariate="detection") + 
  ggtitle("scDD, Covariate 2: Detection Rate")
```

```{r, fig.width=10, fig.height=3.2, message=FALSE}
strat_hist(data.frame(rowData(humandata)), 
           pvalue="scdd_p", covariate="detection", maxy=25) 
strat_hist(data.frame(rowData(humandata2)), 
           pvalue="scdd_p", covariate="detection", maxy=25) 
strat_hist(data.frame(rowData(mousedata)), 
           pvalue="scdd_p", covariate="detection", maxy=25) 
strat_hist(data.frame(rowData(mousedata2)), 
           pvalue="scdd_p", covariate="detection", maxy=25) 
```

#### MAST

```{r, fig.width=4.5, fig.height=3.5, message=FALSE}
rank_scatter(data.frame(rowData(humandata)), 
             pvalue="mast_p", covariate="detection") + 
  ggtitle("MAST, Covariate 2: Detection Rate")
rank_scatter(data.frame(rowData(humandata2)), 
             pvalue="mast_p", covariate="detection") + 
  ggtitle("MAST, Covariate 2: Detection Rate")
rank_scatter(data.frame(rowData(mousedata)), 
             pvalue="mast_p", covariate="detection") + 
  ggtitle("MAST, Covariate 2: Detection Rate")
rank_scatter(data.frame(rowData(mousedata2)), 
             pvalue="mast_p", covariate="detection") + 
  ggtitle("MAST, Covariate 2: Detection Rate")
```

```{r, fig.width=10, fig.height=3.2, message=FALSE}
strat_hist(data.frame(rowData(humandata)), 
           pvalue="mast_p", covariate="detection", maxy=25) 
strat_hist(data.frame(rowData(humandata2)), 
           pvalue="mast_p", covariate="detection", maxy=25) 
strat_hist(data.frame(rowData(mousedata)), 
           pvalue="mast_p", covariate="detection", maxy=25) 
strat_hist(data.frame(rowData(mousedata2)), 
           pvalue="mast_p", covariate="detection", maxy=25) 
```

@@

** Still need to clean up from here on; nothing below is eval'd**

We will skip running the Wilcox method for now.
First we will adjust for multiple comparisons using the scDD results

```{r, benchmarking, eval=FALSE}
